|

### 2019-09-02 paper.Rmd
Small formatting fix
abstract and intro

---
title: |
  A slice tour for finding hollowness in high-dimensional data
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a, b
    email: ursula.laa@monash.edu
  - name: Dianne Cook
    affil: b
    email: dicook@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: |
  Projections and sectioning are complementary methods for the visualisation of
  high-dimensional data. In this note we describe slicing in the orthogonal
  space of projections obtained when running a tour, thus presenting the viewer
  with an interpolated sequence of projected sections. The method has been
  implemented in R as an extension to the tourr package, and can be used to
  explore concave structures in multivariate distributions.
keywords: |
  data visualisation; grand tour; sectioning
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
output: rticles::tf_article
---

@@

<!-- Background -->
Modern data is typically high-dimensional and thus data visualisation often relies on some form of dimension reduction. This entails looking at linear projections of the data, or nonlinear if one considers techniques like multidimensional scaling (MDS) or t-Distributed Stochastic Neighbor Embedding (t-SNE). For our purposes we focus on linear projections. Interactive and dynamic displays can provide information beyond what can be achieved in a static display. For example, the grand tour [@As85; @BCAH05] shows a smooth sequence of interpolated lower dimensional projections and allows the viewer to extrapolate from the low-dimensional shapes to the multidimensional distribution. It is particularly useful for detecting clusters, outliers and non-linear dependence.

<!-- The disadvantage of projection: opacity, occlusion, even with pseudo-transparency from scatterplots;  motivation -->
A major limitation of projections is their opacity, even if mitigated somewhat through the pseudo-transparancy present in sparse scatterplots. As a result certain features of distributions may easily be hidden in projections, especially in the case of concave structures. Considering the example of a simple geometric shape such as a hypershere, we understand that projections cannot distinguish between a full or a hollow sphere. Similarly we can think of small scale structures hidden in the center of a multivariate distribution. Again projections will not help in findig such features in a distribution. Other examples where the use of projections is limited include the visualisation of non-linear boundaries between groups, that will be obscured in each projection.

<!-- Our approach: introduce sections; many different ways to slice, refer to linked brushing as slicing whcih mirrors prosection, but here we will explain/utilise a simple slicing on the projection where point in orthogonal space are faded. -->
We can overcome these limitations when adding section information, see e.g. @prosection. For our example of the hypersphere, we can clearely identify a hollow sphere when looking at a slice throught the distribution, where all points will lie on a circle in a 2d slice. Note that there are many ways to define a slice through high-dimensional data. In practice, slicing is typically facilitated through linked brushing (e.g. @ggobiCSDA2003, @JSSv081i05), where the user is selecting a region in a subspace of the dataspace that is highlighted in a projection of the data points. These systems are interactive, and rely on the user to redefine the section many times to explore the data in this way. In this note we introduce a simple slicing method for projections where points in the orthogonal space are faded. Combining it with the grand tour results in a new dynamic display that can be used to systematically search for features hidden in sections.

<!-- paper structure --> 
The method and its implementation in R are described in Section XXX, and in Section XXX. We then apply it to several examples in Section XXX. We consider different high-dimensional geometric shapes and XXX, showing how concave or occluded structures can be visualised and explored with a slice tour. Possible extensions are discussed in Section XXX.

@@

In general $\alpha$ can be any point in the $n$-dimensional parameter space, but we are only interested in the orthogonal component, thus we use $\alpha' = \alpha - (\alpha\cdot a_1) a_1 - (\alpha\cdot a_2 )a_2$ to define the generalized measure of orthogonal distance as
\begin{equation}
v_i^2 = ||x_i' - \alpha'||^2 = x_i'^2 + \alpha'^2 - 2 x_i'\cdot\alpha'
\end{equation}

where the cross term can be expressed as

\begin{equation}
x_i'\cdot\alpha' = x_i\cdot\alpha - (\alpha\cdot a_1) (x_i\cdot a_1) - (\alpha\cdot a_2) (x_i\cdot a_2).
\end{equation}

>>>>

---
title: |
  A slice tour for finding hollowness in high-dimensional data
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a, b
    email: ursula.laa@monash.edu
  - name: Dianne Cook
    affil: b
    email: dicook@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: |
  Taking projections of high-dimensional data is a common analytical and visualisation technique in statistics for working with high-dimensional problems. Sectioning, or slicing, through high dimensions is less common, but can be useful for visualising data with concavities, or non-linear structure. It is associated with conditional distributions in statistics, and also linked brushing between plots in interactive data visualisation. This short technical note describes a simple approach for  slicing in the orthogonal
  space of projections obtained when running a tour, thus presenting the viewer
  with an interpolated sequence of projected sections. The method has been
  implemented in R as an extension to the tourr package, and can be used to
  explore for concave and non-linear structures in multivariate distributions.
keywords: |
  data visualisation; grand tour; sectioning
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
output: rticles::tf_article
---

@@

<!-- Background -->
Data is commonly high-dimensional, and visualisation often relies on some form of dimension reduction. This can be taking linear projections, or nonlinear if one considers techniques like multidimensional scaling (MDS) or t-Distributed Stochastic Neighbor Embedding (t-SNE). For the purposes here, the focus is on linear projections, in particular as provided by the grand tour [@As85; @BCAH05]. Interactive and dynamic displays can provide information beyond what can be achieved in a static display. The grand tour shows a smooth sequence of interpolated low dimensional projections, and allows the viewer to extrapolate from the low-dimensional shapes to the multidimensional distribution. It is particularly useful for detecting clusters, outliers and non-linear dependence.

<!-- The disadvantage of projection: opacity, occlusion, even with pseudo-transparency from scatterplots;  motivation -->
A major limitation of projections is their opacity. It is mitigated some, when scatterplots are used to render the projected data, through a pseudo-transparancy of sparseness of points. This is improved further if points are also drawn using some transparency. Some features of multivariate distributions may be visible, but a lot is easily hidden, especially in the case of concave structures. Considering the example of a simple geometric shape such as a hypershere, it is difficult to distinguish between a full or a hollow sphere, that is, whether points are uniformly distributed within the sphere or on the surface of a sphere. Similarly we can think of small scale structures hidden in the center of a multivariate distribution, that might be considered to be "needles in a haystack", which can be difficult to detect in projections. Another example is where one would like to examine non-linear boundaries between groups from a classification model, with the nature of the boundaries being occluded in the projections.

<!-- Our approach: introduce sections; many different ways to slice, refer to linked brushing as slicing whcih mirrors prosection, but here we will explain/utilise a simple slicing on the projection where point in orthogonal space are faded. -->
Slicing, or sectioning, is a way that the internal distribution of high-dimensional data can be explored. @prosection discusses a technique for combining projections and sections, by slicing in the back-dimensions. It is also possible to think of linked brushing (e.g. @ggobiCSDA2003, @JSSv081i05) as slicing. In this note we discuss another approach to sectioning, where observations in the space orthogonal to the projection are highlighted if they are close to the projection plane, and faded if further afield. This is combined with the grand tour to provide a new dynamic display that can be used to systematically search for features hidden in high dimensions.

<!-- paper structure --> 
The method is described in Section XXX, and it is implemented in the tourr [@tourr] package in R [@rref].  Section XXX illustrates the use on several high-dimensional geometric shapes and established data sets, showing how concave or occluded structures can be visualised and explored with a slice tour. Future work is discussed in Section XXX.

@@

In general $\alpha$ can be any point in the $n$-dimensional parameter space, but we are only interested in the orthogonal component, thus we use $\alpha' = \alpha - (\alpha\cdot a_1) a_1 - (\alpha\cdot a_2 )a_2$ to define the generalized measure of orthogonal distance as
\begin{equation}
v_i^2 = ||x_i' - \alpha'||^2 = x_i'^2 + \alpha'^2 - 2 x_i'\cdot\alpha'
\end{equation}
where the cross term can be expressed as
\begin{equation}
x_i'\cdot\alpha' = x_i\cdot\alpha - (\alpha\cdot a_1) (x_i\cdot a_1) - (\alpha\cdot a_2) (x_i\cdot a_2).
\end{equation}
